---
title: "Milestone 1"
author: "Angel Waters"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Loading and Cleanup

The necessary packages were loaded for the analysis.
```{r Libraries}
library(ggplot2)
library(tidyverse)
library(ggQC)
```

Load the pre-cleaned up data and subset further.
```{r Load data}
##Loading the DK raw data subset for further cleaning and visualizing the data.
rawData <- 
  read_csv("SWANBaselineData_ProfessorKSubset (1).csv")
str(rawData)

##Further subsets the data to meet the requirements for the final
milestone1_subset <- subset(rawData, select = c(
  SWANID,
  AGE0,
  ANEMIA0,
  SMOKERE0,
  PULSE0,
  HEIGHT0,
  WEIGHT0,
  RACE)
  )
#view the different columns and their structures
view(milestone1_subset)
str(milestone1_subset)

```

Clean up the categorical data for data analysis and presentation.
```{r Data Cleanup}
milestone1_subset <- mutate(milestone1_subset, ANEMIA0=ifelse(ANEMIA0=="(1) No",
                                                              "No", ifelse(ANEMIA0=="(2) Yes", "Yes", NA)),
       SMOKERE0=ifelse(SMOKERE0=="(1) No", "No", ifelse(SMOKERE0=="(2) Yes", "Yes", NA)),
       RACE=ifelse(RACE=="(1) Black/African American", "Black/African American",
                   ifelse(RACE=="(2) Chinese/Chinese American", 
                          "Chinese/Chinese American",
                          ifelse(RACE=="(3) Japanese/Japanese American",
                                 "Japanese/Japanese American",
                                 ifelse(RACE=="(4) Caucasian/White Non-Hispanic", 
                                        "Caucasian/ White Non-Hispanic", 
                                        ifelse(RACE=="(5) Hispanic", "Hispanic", NA)
                                        )))))

```


# Descriptive Statistics

Calculating various descriptive statistics of each of the variables.
```{r Descriptive Stats Numerical}
summary(milestone1_subset)
age_sd <- sd(milestone1_subset$AGE0, na.rm = TRUE)
age_mean <- mean(milestone1_subset$AGE0, na.rm = TRUE)
pulse_sd <- sd(milestone1_subset$PULSE0, na.rm = TRUE)
pulse_mean <- mean(milestone1_subset$PULSE0, na.rm = TRUE)
height_sd <- sd(milestone1_subset$HEIGHT0, na.rm = TRUE)
height_mean <- mean(milestone1_subset$HEIGHT0, na.rm=TRUE)
weight_sd <- sd(milestone1_subset$WEIGHT0, na.rm = TRUE)
weight_mean <- mean(milestone1_subset$WEIGHT0, na.rm=TRUE)
age_med <- median(milestone1_subset$AGE0, na.rm=TRUE)
pulse_med <- median(milestone1_subset$PULSE0, na.rm=TRUE)
height_med <- median(milestone1_subset$HEIGHT0, na.rm=TRUE)
weight_med <- median(milestone1_subset$WEIGHT0, na.rm=TRUE)
```

Create a csv file to use in the report for descriptive statistics.
```{r}
clmn_names <- c("Age", "Pulse", "Height", "Weight")
sds <- c(age_sd, pulse_sd, height_sd, weight_sd)
means <- c(age_mean, pulse_mean, height_mean, weight_mean)
medians <- c(age_med, pulse_med, height_med, weight_med)
desc_table <- data.frame("Continuous Data"=clmn_names, "Standard Deviation"=sds, "Mean"=means, "Medians"=medians)
desc_table
#write the table into a format that can be copied over into the report
write.csv(desc_table, "C:/Users/12072/OneDrive/Desktop/ALY 6010/Final Project\\Milestone1_Descriptive.csv")
```


# Data Visualization

Understand the spread of the numerical data.
```{r Histograms}
age_hist <- ggplot(milestone1_subset)+
  geom_histogram(mapping=aes(AGE0), na.rm = TRUE, binwidth = 1, color="black")+
  theme_classic()+
  labs(title="Spread of Age", x="Age (years)", y="Count")
age_hist
##put a box plot in to highlight the outliers in this chart
pulse_hist <- ggplot(milestone1_subset,mapping=aes(PULSE0), na.rm = TRUE)+
  geom_histogram(color="black")+
  geom_boxplot()+
  theme_classic()+
  labs(title="Spread of Pulse", x="Pulse (beats/30 sec)", y="Count")
pulse_hist
height_hist <- ggplot(milestone1_subset)+
  geom_histogram(mapping=aes(HEIGHT0), na.rm = TRUE, color="black")+
  theme_classic()+
  labs(title="Spread of Height", x="Height (cm)", y="Count")
height_hist
weight_hist <- ggplot(milestone1_subset)+
  geom_histogram(mapping=aes(WEIGHT0), na.rm = TRUE, color="black")+
  theme_classic()+
  labs(title="Spread of Weight", x="Weight (kg)", y="Count")
weight_hist
```

To understand the spread of the categorical data. 
```{r Categorical spread}
##NA's were not being removed with na.rm=TRUE, therefore they were removed before calling the ggplot functions
anem_freq <- milestone1_subset %>% drop_na() %>% 
  ggplot()+
  geom_bar(mapping=aes(ANEMIA0))+
  theme_classic()+
  labs(title= "Frequency of Patients with Anemia", y= "Count", 
       x="Patients has Anemia")
anem_freq  

smoke_freq <- milestone1_subset %>% drop_na() %>% 
  ggplot()+
  geom_bar(mapping=aes(SMOKERE0))+
  theme_classic()+
  labs(title= "Frequency of Patients who Smoke", y= "Count", 
       x="Patients who Smoke")
smoke_freq
##Analysis of RACE
ggplot(milestone1_subset)+
  geom_bar(mapping=aes(RACE))+
  labs(title="Patient Race")+
  theme(axis.text.x=element_text(size=13, angle=45, hjust=1, vjust=1))
```

A separate spread analysis for race was necessary because it has multiple possible results.
```{r Race}
ggplot(milestone1_subset)+
  geom_bar(mapping=aes(RACE))+
  labs(title="Patient Race")+
  theme(axis.text.x=element_text(size=13, angle=45, hjust=1, vjust=1))
##creating frequency and pareto charts to plot the race data
##Creates a tibble with the total counts for each race identified
race <- milestone1_subset %>% group_by(RACE) %>% summarise(Total=n())
##Calculates the frequency of each race
race <- mutate(race, Frequency=Total/sum(Total))
race
race_pareto <- ggplot(race, mapping=aes(x=RACE, y=Total))+
  stat_pareto(bars.fill="gray")+
  labs(title="Patient Race Pareto", x="Race")+
  theme_classic()+
  theme(axis.text.x=element_text(size=13, angle=45, hjust=1, vjust=1))
race_pareto
```

Height and weight relationship using a scatterplot.
```{r Height vs Weight}
hvw_scatter <- ggplot(milestone1_subset)+
  geom_point(mapping=aes(HEIGHT0, WEIGHT0), na.rm=TRUE, size=0.5)+
  geom_smooth(mapping=aes(HEIGHT0, WEIGHT0), method=lm, formula=y~x, na.rm=TRUE)+
  theme_classic()
hvw_scatter
```

Is there a relationship between race and anemia?
```{r Race vs Anemia}
avr_jitter <- milestone1_subset %>% drop_na() %>% ggplot()+
  geom_jitter(mapping=aes(RACE, ANEMIA0), na.rm=TRUE, size=0.5)+
  labs(title="Anemia by Race", x="Race", y="Patient had Anemia Diagnosis?")+
  theme_classic()+
  theme(axis.text.x=element_text(size=10, angle=75, hjust=1, vjust=1))
avr_jitter
```

Is there a relationship between smoking and measured pulse?
```{r smoking vs pulse}
pulse_density <- milestone1_subset %>% drop_na() %>% ggplot()+
  geom_density(mapping=aes(PULSE0, color=SMOKERE0), size=1)+
  theme_classic()+
  labs(title="Density of Pulse by Smoking Status", x= "Pulse (beats/30 sec)", y="Density")
pulse_density
```


