---
title: "Module 4 R Practice"
author: "Angel Waters"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Packages, include=FALSE}
library(pacman)
p_load(MASS)
library(tidyverse)
library(ggplot2)
library(nhstplot)
```


# Cats Body Weights

To understand if there are any differences between the body weights of male cats and female cats, data was analyzed from the R Package MASS to understand statistical differences between the two data sets.

The data was subsetted into two groups based on their sex: male cats and female cats.
```{r Subset Cats Data}
#cats #unquote to view the data
female <- cats %>% filter(Sex=="F")
head(female)
summary(female)
male <- cats %>% filter(Sex=="M")
head(male)
summary(male)
```

Identify the Null Hypothesis, Alternative Hypothesis, and claim.  The claim is that the average body weight of male cats is not equivalent to the average body weight of female cats, and therefore significantly different. This will be analyzed at an $alpha = 0.05$.

$$H_0:\mu_1=\mu_2\\H_1:\mu_1\neq\mu_2$$
```{r Cats Hypotheses and Claim, include=FALSE}
Null <- "Average male body weight is equal to average female body weight"
Alt <- "Average male body weight is not equal to the average female body weight"
Claim <- "The average body weight of male cats is not equivalent to the average body weight of female cats."
```

Calculate the necessary sample statistics for the the t test.
```{r Cats Sample Statistics}
nm <- nrow(male)
nf <- nrow(female)
xbarm <- mean(male$Bwt)
xbarf <- mean(female$Bwt)
sm <- sd(male$Bwt)
sf <- sd(female$Bwt)
alpha <- 0.05
dfm <- nm-1
dff <- nf-1
df <- min(dfm, dff) #df to be used 
```

This is a two tailed test because it is trying to disprove equivalence with non-equivalence, and not focus on greater than or lesser than results.  The critical values were calculated and plotted.
```{r Cats critical values}
halpha <- alpha/2
cv <- qt(p=1-halpha, lower.tail = TRUE, df=df)
cv
#draw curve set to default as a two tailed t test
plotttest(cv, df=df)
```

t test was used to calculate the significance between the two sample averages.
```{r Cats t test}
##Manual calculation of the t statistic
t <- ((xbarm-xbarf)-(0))/sqrt(((sm^2)/nm)+((sf^2)/nf))
t
##using the R tool to calculate the t statistic
tsummary <- t.test(female$Bwt, male$Bwt, alternative="two.sided", var.equal = FALSE, mu=0, conf.level = 0.95)
tsummary
```

Based on the summary statistics, the decision around the hypothesis test is:
```{r Cats decision, echo=FALSE}
decision <- if(abs(t)-abs(cv)>0){
  "Reject the Null Hypothesis"
} else {
  "Do not reject the Null Hypothesis"
}
decision
```

## Conclusion
Based on the analysis presented above the following results were summarized about the differences in body weight between male and female cats:
```{r Cats conclusion, echo=FALSE}
conclusion <- if(decision=="Reject the Null Hypothesis"){
  "There is enough evidence to support the claim:"
} else{
  "There is not enough evidence to support the claim:"
}
cat(conclusion, Claim)
```

\newpage

# Meditation Effects on Sleep Quality

A meditation seminar was given to 10 students to see if the workshop improved sleep quality for these students.  A paired student's t test was used to analyze if the difference in sleep quality before and after the meditaion workshop was significant and therefore the quality of sleep for the student's improved.

```{r Meditation Dataset}
before <- c(4.6,7.8,9.1,5.6,6.9,8.5,5.3,7.1,3.2,4.4)
after <- c(6.6,7.7,9.0,6.2,7.8,8.3,5.9,6.5,5.8,4.9)
```

Identify the Null Hypothesis, Alternative Hypothesis, and claim.  The claim is that the sleep quality of the students improved after they took the meditation workshop. This will be analyzed at an $alpha = 0.05$.

$$H_0:\mu_D=0\\H_1:\mu_D<0$$

```{r Meditation Hypotheses and claims, include=FALSE}
Null <- "uD = 0"
Alt <- "uD < 0"
Claim <- "Meditation increases the average sleep score"
```

The initial step for gathering the necessary sample statistics is to calculate the individual differences for each pair of points.
```{r Differences, echo=FALSE}
Meditation <- data.frame(before, after)
Meditation <- Meditation %>% mutate(D = before-after, D2= D^2)
Meditation
```

This is a left tailed t test which means the critical value will be calculated via the one-tailed method.
```{r Meditation critical values}
n <- nrow(Meditation)
alpha <- 0.05
df <- n-1
cv <- qt(p=1-alpha, df=df, lower.tail = FALSE )
```

The summary statistics were then calculated for the t test.
```{r Meditation t.test calculation}
dbar <- mean(Meditation$D)
sumd <- sum(Meditation$D)
sumd2 <- sum(Meditation$D2)
sD <- sqrt(((n*sumd2)-(sumd)^2)/(n*(n-1)))
t <- dbar/(sD/sqrt(n))
```

Based on the t statistic, the following decision was made about the sleep scores:
```{r Meditation Decision, echo=FALSE}
decision <- if(abs(t)-abs(cv)>0){
  "Reject the Null Hypothesis"
} else {
  "Do not reject the Null Hypothesis"
}
decision
```

## Conclusion

Based on the analysis above, the conclusion can be made around the claim:
```{r Meditation Conclusion, echo=FALSE}
conclusion <- if(decision=="Reject the Null Hypothesis"){
  "There is enough evidence to support the claim:"
} else{
  "There is not enough evidence to support the claim:"
}
cat(conclusion, Claim)
```

The method for analyzing the meditation dataset was the two-sample paired t test.  This was identified as the appropriate mode for analysis because the data was a before and after study.  Each point collected was dependent on a matched point, the student's sleep quality analyzed before the meditation class and the same student's sleep quality analyzed after the meditation class.  The aim of the study is to look at the difference therefore making the samples matched-pairs.
